---
title: 'EGU short course: 1. Spatiotemporal change detection'
author: "Meng Lu"
date: "April 18, 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE , tidy.opts=list(width.cutoff = 60), tidy = TRUE, warning = FALSE)
```
## Method: 
Integrate SAR (simutaneous autocorrelation regression) to the empirical fluctutaion process (efp) structural change test. 

## What to do in the next 15 minutes: 
* Seasonality analysis of a time series
* efp (empirical fluctuation process, from the R package "strucchange") and BFAST (breaks for additive seasonality and trend, from the R package "BFAST") methods for time series structural change detection.
* Spatial correlation of the area
* SAR integrated efp 

## Start!
   
```{r loadpackages, include = FALSE}
 ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE, repos = "https://cran.uni-muenster.de/")
  sapply(pkg, require, character.only = TRUE)
}
packages <- c("strucchange", "spacetime", "spdep", "nlme", "zoo", "gstat", "spacetime", "sp")
ipak(packages)
 
lapply(packages, require, character.only = TRUE)
```

Load data, "fevi8" is a 3d array with longitude, latitude, and time as dimensions. 
```{r}
load("Rdata/fevi8.Rdata")
dim(fevi8)
```



### Time series structural change analysis
#### 1. BFAST (breakpoints for additive seasonality and trend): detecting change in seasonality and trend interatively

Choose a location and form a time series from the data matrix
```{r}
lon = 60
lat = 40
originalts <- ts(fevi8[lon, lat, ], start=c(2000, 1), frequency=46)
```
Assuming additive seasonality and trends, use stl (loess) to separate trend, seasonality and residuals. 

```{r}
seasonality <- stl(originalts, s.window = "per")$time.series[,"seasonal"]
plot(stl(originalts, s.window = "per"))
#spec.ar(seasonality)
le = 636 # the length of time series
tl = 1:le
```

##### remove seasonality and detect change in trend
```{r}
trend_rmstlsea <- originalts - seasonality
```

Use the empirical fluctuation test to test structural change in trend. The red lines indicate threshold of a change. 

```{r}
efp_trend <- efp(trend_rmstlsea ~ tl, type = "OLS-MOSUM")
plot(efp_trend)
sctest(efp_trend)
```

##### remove trend and detect change in seasonality
```{r}
seasonality <- originalts -  fitted(lm(trend_rmstlsea ~ tl)) 
plot(seasonality)
```

remove trend:
seasonality = originalts - trend
Form harmonic terms. 
 
$x_t = Acos(2\pi wt + \phi)$

$x_t = U_1 cos(2\pi wt) + U_2 sin(2\pi wt)$

($cos(\alpha + \beta) = cos(\alpha)cos(\beta) - sin (\alpha)sin(\beta)$)
```{r harmonic model}
w = 1 / 46
#harmonic
co <- cos(2 * pi * tl * w)
si <- sin(2 * pi * tl * w)
co2 <- cos(2 * pi * tl * w * 2)
si2 <- sin(2 * pi * tl * w * 2)  
co3 <- cos(2 * pi * tl * w * 3)
si3 <- sin(2 * pi * tl * w * 3) 
```


Fit a first order harmonic model to the seasonality and model the seasonality in residuals.
```{r}
res_har1 <- residuals(lm(seasonality ~ co + si ))
summary(lm(seasonality ~ co + si))
 
#res_har1 <- ts(res_har1, start=c(2000, 1), frequency=46)
#sea_har1 <- stl(res_har1, s.window = "per")$time.series[,"seasonal"]
#plot(stl(res_har1, s.window = "per"))
#spec.ar(res_har1)
```

Fit a second order harmonic model to the seasonality and model the seasonality in residuals. 

```{r}
res_har2 <- residuals(lm(seasonality ~ co + si +co2 + si2))
summary(lm(seasonality ~ co + si + co2 + si2))

#res_har2 <- ts(res_har2, start=c(2000, 1), frequency = 46)
#sea_har2 <- stl(res_har2, s.window = "per")$time.series[,"seasonal"]
#plot(stl(res_har2, s.window = "per"))
#spec.ar(res_har2)
```

As the third order of harmonics explain more variance, we will use the third order harmonics.
```{r}
res_har3 <- residuals(lm(seasonality ~ co + si +co2 + si2 + co3 + si3))
summary(lm(seasonality ~ co + si +co2 + si2 + co3 + si3))
#res_har3 <- ts(res_har3, start=c(2000, 1), frequency=46)
#sea_har3<- stl(res_har3, s.window = "per")$time.series[,"seasonal"]
#plot(stl(res_har3, s.window = "per"))
#spec.ar(sea_har3)
```

Empirical fluctuation test of change in seasonality

```{r}
p.Vt1 <-  efp(seasonality ~ co + co2 + co3 + si + si2 + si3, h = 0.15, type = "OLS-CUSUM")
plot(p.Vt1) 
```


#### 2. Regression on trend and harmonic terms at once: the BFAST Monitor method: 
```{r}
    p.Vt1 <-  efp(originalts ~ tl + co + co2 + co3 + si + si2 + si3, h = 0.15, type = "OLS-CUSUM")
    plot(p.Vt1) 
    p.Vt1 <-  efp(originalts ~ tl + co + co2 + co3 + si + si2 + si3, h = 0.15, type = "OLS-MOSUM")
    plot(p.Vt1) 
```

 
### Spatial correlation
Here we checked a parameter (e.g. cosine coefficient) of seasonality. We could also check the spatial correlation in seasonality of model (e.g. BFAST) residuals, trend coefficients, as well as other seasonality coefficients. 
```{r, eval = FALSE}
coef_sea <- function(ts1, whichcoef)
{ 
  originalts <- ts(ts1, start=c(2000, 1), frequency=46)
  seasonality <- stl(originalts, s.window = "per")$time.series[, "seasonal"]
  trend = originalts - seasonality 
  coefficients(lm(seasonality ~ co + si))[whichcoef]
}
 seacoefco <- apply(fevi8, c(1, 2), coef_sea, 2) # coefficients of the consine term
 seacoefsi <- apply(fevi8, c(1, 2), coef_sea, 3) # coefficients of the sine term
 save(seacoefsi, file = "seacoefsi.Rdata")
 save(seacoefco, file = "seacoefco.Rdata")
```

The seasonality coefficient takes around 1 min to run. In case of saving some time, we can load the seasonality coefficeints.  

```{r}
load("Rdata/seacoefsi.Rdata")
load("Rdata/seacoefco.Rdata")
```

We could see a spatial pattern in the seasonality coefficient: the sine term

```{r}
coor <- expand.grid(x = 1:dim (fevi8)[1], y = 1:dim (fevi8)[2])
sdfsi <- data.frame(seacoef = as.vector(seacoefsi), coor)  
coordinates(sdfsi) <- ~x + y
sdf1 <- sdfsi
gridded(sdf1) <- TRUE
spplot(sdf1)
```
We could see a spatial pattern in the seasonality coefficient: the cosine term
```{r}
sdfco <- data.frame(seacoef = as.vector(seacoefco), coor)  
coordinates(sdfco) <- ~x + y
sdf1 <- sdfco
gridded(sdf1) <- TRUE
spplot(sdf1)
```

We could also have a look at the variogram and fit a variogram model. Here I regressed on locations to get rid of spatial trend. It is clear that semivariance increase with distance, which indicates spatial correlation.
The sine term

```{r}
v <- variogram(seacoef ~ x + y, sdfsi)
model3d <- fit.variogram(v, vgm( c("Exp", "Ste", "Sph", "Mat", "Gau") )) 
model3d
plot(v, model3d)
```
The cosine term:
```{r}
v <- variogram(seacoef ~ x + y, sdfco)
model3d <- fit.variogram(v, vgm( c("Exp", "Ste", "Sph", "Mat", "Gau") )) 
model3d
plot(v, model3d)
```

#### SAR integrated efp model:

Create spatiotemporal cubes and weight matrix.

```{r}
  eday <- as.Date("2000-01-30")           # date 
  e8day <- seq(eday, length.out = 636, by = "8 days")
  xyd <- expand.grid(x1 = 1:3, y1 = 1:3)
  coordinates(xyd) <- ~x1 + y1
  lecube <- 3 * 3 * 636
  aa3 <- as.data.frame(c(1:lecube))
  stfdf3b3 <- STFDF(xyd, e8day, aa3) ## for creating neighbors only, aa3 could be any data
  cn <- cell2nb(3, 3, type = "queen", torus = FALSE)
  neigh1 <- nbMult(cn, stfdf3b3, addT = FALSE, addST = FALSE) # only spatial neighbours are added for each time step
  listcn636 <- nb2listw(neigh1)
```

Regressors (trend and seasonality) in a matrix

```{r}
 X = matrix(0, 636 * 9, 9 * 8)
  
  for( i in 1:9)
  {
    X[seq(i, by = 9, length.out = 636), 1 + (i - 1) * 8] = 1 
    X[seq(i, by = 9, length.out = 636), 2 + (i - 1) * 8] = tl 
    X[seq(i, by = 9, length.out = 636), 3 + (i - 1) * 8] = co
    X[seq(i, by = 9, length.out = 636), 4 + (i - 1) * 8] = co2
    X[seq(i, by = 9, length.out = 636), 5 + (i - 1) * 8] = co3
    X[seq(i, by = 9, length.out = 636), 6 + (i - 1) * 8] = si
    X[seq(i, by = 9, length.out = 636), 7 + (i - 1) * 8] = si2
    X[seq(i, by = 9, length.out = 636), 8 + (i - 1) * 8] = si3
    colnames(X) = paste0("v", 1:(9 * 8))
    X
      }
```

Load the modified version of strucchange, the only difference is the change in the efp function, for OLS-MOSUM and OLS-CUSUM tests. In the modified verson, structural change is analysed directly from the residuals of spatialtemporal model. The function efp() takes a "spatial1" variable (i.e. the modified version of efp: efp <- function(....,  spatial1 = list()))
and skip the linear regression formula. The "spatial1" contains a list of residuals from SAR integrated time series regression mode.

```{r}
library(devtools)
install_github("mengluchu/strucchange",build_vignettes = FALSE)
```

SAR integrated efp. The most time-consuming process is the SAR model (spautolm). It costs 22 seconds to run on my computer.

```{r}
      f2 <- fevi8[(lon-1):(lon + 1), (lat-1):(lat + 1), ]
      fevi3b312t1 <- ts(f2[2, 2, ], start = c(2000, 1), frequency = 46) # reconstruct the time series
      aa2 <- as.vector(f2) 
      system.time(try2 <- spautolm(aa2~. , data.frame(aa2, X),family = "SAR", method = "Matrix", listw = listcn636))   
      rn <- lapply(1:9,function(i) {residuals(try2)[seq(i, 636 * 9 - (9 - i), 9)]})
      p.Vt1 <- efp(fevi3b312t1 ~ 1, h = 0.15, type = "OLS-CUSUM", spatial1 = as.numeric(rn[[5]]))
      plot(p.Vt1)
      sctest(p.Vt1)$p.value
      
```

OLS-MOSUM method:

```{r}
p.Vt1 <- efp(fevi3b312t1 ~ 1, h = 0.15, type = "OLS-MOSUM", spatial1 = as.numeric(rn[[5]]))
      plot(p.Vt1)
```

Comparing with the pure time series anaylsis: OLS-CUSUM
```{r}
p.Vt1 <- efp(fevi3b312t1 ~tl + co + co2 + co3 + si + si2 + si3, h = 0.15, type = "OLS-CUSUM")
      plot(p.Vt1)
```
OLS-MOSUM
```{r}
p.Vt1 <- efp(fevi3b312t1 ~tl + co + co2 + co3 + si + si2 + si3, h = 0.15, type = "OLS-MOSUM" )
      plot(p.Vt1)
```
     

Detect change using structural change test and store the p-value into an array. Here is an example conduct the analysis for 4 $3*3*636$ spatiotemporal cubes.    


```{r, eval= FALSE} 
 tssar1 <- array(NA , c(2, 2))
 
   for(i in 30:31)
  {
    for (j in 30:31 )
    {
      f2 <- fevi8[i:(i + 2), j:(j + 2), ]
      fevi3b312t1 <- ts(f2[2, 2, ], start = c(2000, 1), frequency = 46) # reconstruct the time series
      aa2 <- as.vector(f2) 
      try2 <- spautolm(aa2~. , data.frame(aa2, X),family = "SAR", method = "Matrix", listw = listcn636)   
      rn <- lapply(1:9,function(i) {residuals(try2)[seq(i, 636 * 9 - (9 - i), 9)]})
      p.Vt1 <- sctest(efp(fevi3b312t1 ~ 1, h = 0.15, type = "OLS-CUSUM", spatial1 = as.numeric(rn[[5]])))
      tssar1[i, j] < -p.Vt1$p.value 
    } 
  } 
```
Pure time series analysis:
```{r, eval = FALSE} 
  ts1 <- array(NA , c(2, 2))
  system.time(
    for(i in 1:2)
   {
    for (j in 1:2)
    {
      f2 <- fevi8[i:(i + 2),j:(j + 2), ]
      fevi3b312t1 <- ts(f2[2, 2, ], start = c(2000, 1), frequency = 46) # reconstruct the time series
      p.Vt1 <- sctest(efp(fevi3b312t1 ~ tl + co + co2 + co3 + si + si2 + si3, h = 0.15, type = "OLS-CUSUM"))
      ts1[i, j] <- p.Vt1$p.value 
    } 
  })
```

P-values for each pixels.
```{r, eval = FALSE}
tssar1
ts1
```


### Scale the SAR-efp with SciDB and reproduce the results of a study case in "Spatio-Temporal Change Detection from Multidimensional Arrays: detecting deforestation from MODIS time series", ISPRS journal, Mar, 2016:
https://github.com/mengluchu/scalable-spatial-temporal-BFAST


